---
title: "Vaccine Effectiveness in Catalunya"
author: "Johannes Burr"
date: "16 12 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
We'll write some introductory words later.

## Data Preparation

Since this is a real data set it naturally does not come precisely in the form our analysis expects it to be. Therefore, we first have to prepare and also translate it.

```{r}
data = read.csv("Impacte_del_COVID.csv", encoding="UTF-8")
df = data
colnames(df) = c("sex", "age", "date", "event", "vaccinated", "count")

df$sex = as.factor(df$sex)               # male, female, NaN
df$age = as.factor(df$age)               # discretizied for some reason
df$event = as.factor(df$event)           # positive, hospitalized, critical (ICU)
df$date = as.Date(df$date,"%d/%m/%Y")    # now in format Year,month,day
df$vaccinated = as.factor(df$vaccinated) # none, partial, full
df$count = as.numeric(df$count)  # this means #count people with these characteristics have been observed on this day

# the task asks us to only analyze people older than 30
df = subset(df, age=="30 a 39" | age == "40 a 49" | age == "50 a 59" |
         age == "60 a 69" | age == "70 a 79" | age == "80 o més")

# also only between 1st March 2020 and 12. December 2021
df = subset(df, date>"2021-03-01" | date > "2021-12-12")
dim(df)
```
I have pasted the task from the pdf below. We should at least do that.

## Descriptive Analysis
### Plots
Histograms, Bar charts, Pareto Plots.

### Critical Patients over time
divided by gender (and maybe also vaccination status?)

### Contingency Table
Count as function of vaccination status and age

## Statistical Inference
Poisson Model to fit this count variable

## Conclusion
